QuickSched: Task-based parallelism with dependencies and conflicts

### 2014-09-21 paper/paper.tex
yet another iteration.

% Title portion
\title{QuickSched: Task-based parallelism with dependencies and conflicts}
\author[ecs,ggl]{Pedro Gonnet\corref{cor}} \ead{pedro.gonnet@durham.ac.uk}
\author[ecs]{Aidan B. G. Chalk} \ead{aidan.chalk@durham.ac.uk}
\author[icc]{Matthieu Schaller} \ead{matthieu.schaller@durham.ac.uk}
\cortext[cor]{Corresponding author}
\address[ecs]{School of Engineering and Computing Sciences, Durham University, South Road, Durham DH1 3LE, United Kingdom}
\address[icc]{Institute for Computational Cosmology, Durham University, South Road, Durham DH1 3LE, United Kingdom}
\address[ggl]{Google Switzerland GmbH, Brandshenkestr. 110, 8002 Z\"urich, Switzerland}

@@

Task-based parallelism is a conceptually simple paradigm for
shared-memory parallelism in which a computation is broken-down
into a set of inter-dependent tasks which are executed
concurrently.
Task dependencies are used to model the flow of data between
tasks, e.g.~if task $B$ requires some data generated by task $A$,
then task $B$ {\em depends} on task $A$ and cannot be executed
before task $A$ has completed.
The tasks and their dependencies can be seen as the nodes and edges
respectively of a Directed Acyclic Graph (DAG) which can be
traversed in topological order, executing the tasks at the nodes
on the way down.

@@

In SMP superscalar \cite{ref:Perez2008}, StarPU \cite{ref:Augonnet2011},
QUARK \cite{ref:Yarkhan2011}, and KAAPI \cite{ref:Gautier2007}
the programmer specifies
what shared data each task will access, and how that data will
be accessed, e.g.~read, write, or read-write access.
The dependencies between tasks are then generated
automatically by the runtime system, assuming that the
data must be accessed and updated in the order in which
the tasks are generated.
StarPU also provides an interface for specifying additional
dependencies explicitly.
Intel's Threading Building Blocks (TBB)
\cite{ref:Reinders2010}
provide task-based parallelism using C++ templates.
Dependencies are handled either by explicitly waiting
for spawned tasks, or by explicitly manipulating
task reference counters.

@@

In all of these systems, the tasks are only aware of a single
type of relationship between each other, i.e. dependencies, which
specify a strict ordering between two tasks.
In many cases, however, the task ordering need not necessarily
be this strict.
Consider the case of two tasks that update some shared resource
in an order-independent way, e.g. when accumulating a result in
a shared variable, or exclusively writing to an output file.
In order to avoid concurrent access to that resource, it is
imperative that the execution of both tasks do not overlap,
yet the order in which the tasks are executed is irrelevant.
In the following, such a relationship will be referred to
as a ``conflict'' between two tasks.
\fig{TaskConflicts} shows a task graph with conflicting tasks
joined by thick dashed lines.
None of tasks $F$, $H$, and $I$ can be executed concurrently,
i.e. they must be serialized, yet in no particular order.

@@

Conflicts can be modeled directly as exclusive locks on a shared resource
which have to be obtained by a task before it can execute.
Thus, in \fig{TaskConflicts}, before executing, task $F$ has
to obtain an exclusive lock on the resource associated with
the conflict between tasks $F$, $H$, and $I$.
While task $F$ is being executed, neither $H$ nor $I$ can 
lock the same resource, and therefore will not execute until
task $F$ is done and the lock has been released.

In order to be of practical use, and to scale well with modern
multi-core shared-memory architectures, any task scheduler
implementing constraints must be:
\begin{itemize}
    \item {\em Correct}: All constraints, i.e.~dependencies and
        conflicts, must be correctly enforced,
    \item {\em Fast}: The overheads associated with task management
        should be as small as possible,
    \item {\em Memory/cache-efficient}: Tasks accessing similar
        sets of data should be preferentially executed on the
        same core to preserve memory/cache locality as far as possible, and
    \item {\em Parallel-efficient}: Tasks should be executed in an order
        that sufficient work is available for all computational
        threads at all times.
\end{itemize}
\noindent This paper presents QuickSched, a framework for task-based
parallel programming with constraints, which aims to achieve
all of these goals.
Section~2 describes the main design considerations, Section~3 the
underlying algorithms and data structures, and
Section~4 their specific implementation in QuickSched.
Section~5 presents two test-cases:
\begin{enumerate}
    \item The tiled QR
    decomposition described in \cite{ref:Buttari2009} and for
    which the QUARK scheduler was originally developed, and
    \item A task-based Barnes-Hut tree-code to compute the
    gravitational N-body problem similar to the FMM codes
    of \cite{ref:Ltaief2012} and \cite{ref:Agullo2013},
\end{enumerate}
These real-world examples show how QuickSched can be used in practice,
and can be used to assess its efficiency.
Section~6 concludes with some general observations and future work
directions.

@@

From a programmer's perspective, there are two main paradigms for generating
task dependencies:
\begin{itemize}
  \item Implicitly via spawning and waiting, e.g.~as in Cilk, or
  \item Automatic extraction from data dependencies.
\end{itemize}

The first scheme, spawning and waiting, is arguably the simplest to
use.
For simple depedency structures in which each task depends on only a
single task, i.e.~if the task DAG is a tree, each task {\em spawns}, or
creates, its dependent tasks after its completion.
Hence for the tasks $A$--$E$ in \fig{Tasks}, task $A$ would spawn
tasks $B$ and $D$, task $B$ would spawn task $C$, and task $D$ would
spawn task $E$.
If a task has more than one dependency, e.g. tasks $D$--$F$ in \fig{Tasks},
then the task generation order is reversed: Task $E$ is executed first,
and first spawns tasks $D$ and $F$, and waits for both their completion
before doing its own computations.

@@

Automatic dependency extraction works by enforcing dependencies
between tasks that access the same data.
These data dependencies are provided explicitly by the programmer, e.g.~by
describing which parameters to a task are input, output, and input/output.
The dependencies are enforced in the order in which the tasks are 
created.
This approach usually relies on compiler extensions, e.g.~ {\tt pragma}s
in C/C++, or a system of runtime wrappers, to describe the task parameters
and their intent.

This approach allows programmers to specify rich dependency hierarchies
with very little effort, they still only allow for one type of relationship,
i.e.~dependencies, and lack the ability to deal with conflicts as
described in the previous section.

Here, QuickSched takes a drastically different approach.
The programmer creates the task DAG explicitly before execution.

@@

The division of labor regarding {\em correctness}
between the scheduler and
the queue objects is illustrated in \fig{QSched}.
The scheduler holds the tasks and is in charge
of managing {\em dependencies}.
Once a task has no unresolved dependencies, it is passed
on to a queue object.
The queue object, on the other hand, is in charge
of managing {\em conflicts}.
Computational threads can query a queue and will
receive only tasks for which all conflicts have been
resolved, i.e. for which all necessary resources could be 
exclusively locked.

There is also a division of responsibilities regarding
{\em efficiency} between the scheduler and the queue
objects.
The tasks in each queue are grouped according to the resources
they use, i.e. all the tasks in the same queue use a
similar set of resources.
The underlying assumption is that each computational
thread will preferentially access the same queue for tasks.
If the tasks in the queue share the same set of resources,
in increases the probability of said resources already
being present in the thread's cache, thus increasing
{\em memory efficiency}.
The scheduler is in charge of selecting the most appropriate
queue for each task, based on information stored in each task
on which resources are used.
Given a set of tasks with similar resources for which all
dependencies are resolved, the queue then decides which
tasks to prioritize.
This decision is made based on the length of the critical
path of the dependencies of each task.

@@

The {\tt unlocks} field points to the first element of
an array of {\tt nr\_unlocks} pointers to other tasks.
These pointers represent the dependencies in reverse:
if task $B$ depends on task $A$, then task $A$ {\em unlocks}
task $B$.
The unlocks therefore follow the direction of the arrows
in \figs{Tasks}{TaskConflicts}.
Conversely, {\tt wait} is the number of unresolved dependencies
associated with this task, i.e.~the number of unexecuted tasks
that unlock this task.
Given an array of {\tt N} tasks, the wait-counters can
be set as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
int j, k;
for ( k = 0 ; k < N ; k++ )
  tasks[k].wait = 0;
for ( k = 0 ; k < N ; k++ )
  for ( j = 0 ; j < tasks[k].nr_unlocks ; j++ )
    tasks[k].unlocks[j]->wait += 1;
    \end{lstlisting}
\end{minipage}\end{center}

@@

Finally, {\tt cost} and {\tt weight} are measures
for the relative computational cost of this task, and the
relative cost of the critical path following the
dependencies of this task, respectively.
The task weights can be computed by first sorting
the tasks topologically according to their dependencies, i.e.
as per \cite{ref:Kahn1962}:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
int top[N], j, k, sorted = 0, maxweight;
set the task waits as described above.
for ( k = 0 ; k < N ; k++ )
  if ( tasks[k].wait == 0 )
    top[ sorted++ ] = k;
for ( k = 0 ; k < sorted ; k++ )
  for ( j = 0 ; j < tasks[ top[k] ].nr_unlocks ; j++ )
    if ( ( tasks[ top[k] ].unlock[j]->wait -= 1 ) == 0 )
      top[ sorted++ ] = k;
if ( k < N )
  error( "Circular dependencies detected." );
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the array {\tt top} contains the task indices
in topological order. 
The test in line~10 is a convenient check if the tasks and their
dependencies actually do form an acyclic graph.
The weights themselves are then computed as follows
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
for ( k = N-1 ; k >= 0 ; k-- ) {
  for ( w = 0 , j = 0 ; j < tasks[ top[k] ].nr_unlocks ; j++ ) {
    tasks[ top[k] ].unlock[j]->wait += 1;
    w = max( tasks[ top[k] ].unlock[j]->weight , w );
    }
  tasks[ top[k] ].weight = tasks[ top[k] ].cost + w;
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the tasks are traversed in reverse
topological order, computing the recursive weight as the sum of the
task cost and the maximum weight of the tasks it unlocks (line~6),
and recomputing the task waits at the same time (line~3).

@@

The {\tt lock} field is either {\tt 0} or {\tt 1} and indicates
whether this resource is currently in use, i.e.~{\em locked}.
To avoid race conditions, this value should only be tested
and set using atomic instructions.
The {\tt hold} field is a counter indicating how many
sub-resources of the current resource are locked.
If a resource's hold counter is not zero, then it is
{\em held} and cannot be locked.
Likewise, if a resource is locked, it cannot be held
(see \fig{Resources}).

@@

Incrementing the hold counter of the resource can be implemented
as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void resource_hold ( struct resource *r ) {
  if ( atomic_cas( &r->lock , 0 , 1 ) != 0 )
    return 0;
  atomic_inc( &r->hold );
  r->lock = 0;
  return 1;
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt atomic\_cas(val,old,new)} is an atomic
compare-and-swap operation that sets {\tt val} to {\tt new}
if it is currently equal to {\tt old}.
Similarly, {\tt atomic\_inc(val)} increments {\tt val} by one
atomically.
The resource's {\tt lock} is used to check if the resource
is already locked (line 2), and it is held while {\tt hold}
is incremented, to avoid overlapping hold/lock operations.
If the resource can be locked, the hold counter is incremented
atomically (line~4), and the lock is released (line~5),
returning {\tt 1} or {\tt 0} if the resource could be held
or not, respectively.

The locking procedure itself is implemented as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void resource_lock ( struct resource *r ) {
  struct resource *up, *top;
  if ( r->hold && atomic_cas( &r->lock , 0 , 1 ) != 0 )
    return 0;
  if ( r->hold ) {
    r->lock = 0;
    return 0;
    }
  for ( up = r->parent ; up != NULL ; up = up->parent ) {
    if ( !resource_hold( up ) )
      break;
  if ( ( top = up ) != NULL ) {
    for ( up = r->parent ; up != top , up = up->parent )
      atomic_dec( &up->hold );
    r->lock = 0;
    return 0;
    }
  else
    return 1;
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where in line~3 the resource is first locked if it
is not held.
Due to a possible race condition when holding the resource,
the {\tt hold} counter must checked again once the resource
has been locked (line~5).
In lines~9--11 the hold counters of the hierarchical parents
are incremented using the procedure described earlier.
If this process fails at any point (line~12), the
previously set hold counters are decremented (lines~13--14)
and the lock is released (line~15).
The procedure then returns {\tt 1} or {\tt 0} if the resource
could be locked or not, respectively.

Finally, unlocking a resource is relatively straight-forward:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void resource_unlock ( struct resource *r ) {
  struct resource *up;
  r->lock = 0;
  for ( up = r->parent ; up != NULL ; up = up->parent ) {
    atomic_dec( &up->hold );
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the resource itself is unlocked (line~3)
and the hold counters of all of its hierarchical parents
are decremented (lines~4--5).

@@

One possible strategy would be to maintain an array of tasks
sorted by their weights, and to traverse that list in descending
order, trying to lock the resources of each task, until
a lockable task is found, or returning a failure otherwise.
Although this would return the best possible task, it
requires maintaining a sorted list in which inserting
or removing an entry is in \oh{n} for $n$ elements.
Using an unsorted array would require only \oh{1} operations for
insertion and deletion, but is undesirable as it completely
ignores the task weights.

@@

The data structure for the queue is thus defined as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct queue {
  struct task **tasks;
  int count, lock;
  };
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt tasks} is an array of {\tt count} pointers
to the tasks in max-heap order, and {\tt lock} is used to
guarantee exclusive access to the queue.

Inserting a task in the queue is relatively straight-forward:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void queue_put ( struct queue *q , struct task *t ) {
  while ( atomic_cas( q->lock , 0 , 1 ) != 0 );
  q->tasks[ q->count++ ] = t;
  bubble-up the q->count-1st entry of q->tasks.
  q->lock = 0;
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the loop in line~2 spins until an exclusive
lock on the queue can be obtained.
The task is added to the end of the heap array (line~3)
and the heap order is fixed (line~4).
Before returning, the lock on the queue is released (line~5).

Obtaining a task from the queue can be implemented as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *queue_get ( struct queue *q ) {
  struct task *res = NULL;
  int j, k;
  while ( atomic_cas( q->lock , 0 , 1 ) != 0 );
  for ( k = 0 ; k < q->count ; k++ ) {
    for ( j = 0 ; j < q->tasks[k]->nr_locks ; j++ )
      if ( !resource_lock( q->tasks[k]->lock[j] ) )
        break;
    if ( j < q->tasks[k]->nr_locks )
      for ( j = j-1 ; j >= 0 ; j-- )
        resource_unlock( q->tasks[k]->lock[j] );
    else
      break;
    }
  if ( k < q->count ) {
    res = q->tasks[k];
    q->tasks[k] = q->tasks[ --q->count ];
    trickle-down the kth entry of q->tasks.
    }
  q->lock = 0;
  return res;
  }
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where as with the queue insertion, the queue is first
locked for exclusive access (line~4).
The array of task pointers is then traversed in heap order (line~5),
locking the resources of each task (lines~6--8).
If any of these locks fail (line~9), the locks that were obtained
are released (lines~10--11), otherwise, the traversal is aborted
(line~13).
If all the locks on a task could be obtained (line~15), the
task pointer is replaced by the last pointer in the heap (line~17)
and the heap order is restored (line~18).
Finally, the queue lock is released (line~19) and the locked task
or, if no lockable task could be found, {\tt NULL} is returned.

@@

This section presents two test cases showing
how QuickSched can be used in real-world applications, and
providing benchmarks to assess its efficiency and scalability.

The first test is the tiled QR decomposition originally
described in \cite{ref:Buttari2009}, which has been used as a benchmark
by other authors \cite{ref:Agullo2009b,ref:Badia2009,ref:Bosilca2012}.
This example only requires dependencies and is presented 
as a point of comparison to existing task-based parallel
programming infrastructures.

The second example is a Barnes-Hut tree-code, a problem
similar to the Fast Multipole Method described in both
\cite{ref:Ltaief2012} and \cite{ref:Agullo2013}.
This example shows how conflicts, modeled
via hierarchical resources, can be useful.

The source code of both examples is distributed with the
QuickSched library, along with scripts to run the benchmarks
and generate the plots used in the following.
All examples were compiled with gcc v.\,4.8.1 using the
{\tt -O2 -march=native} flags and run on a 
tests were run on a 64-core AMD Opteron 6376 machine running
at 2.6\,GHz.

>>>>

% Title portion
\title{QuickSched: Task-based parallelism with dependencies and conflicts}
\author[ecs,ggl]{Pedro Gonnet\corref{cor}} \ead{pedro.gonnet@durham.ac.uk}
\author[ecs]{Aidan B. G. Chalk} \ead{aidan.chalk@durham.ac.uk}
\author[icc]{Matthieu Schaller} \ead{matthieu.schaller@durham.ac.uk}
\cortext[cor]{Corresponding author}
\address[ecs]{School of Engineering and Computing Sciences, Durham University, South Road, Durham DH1 3LE, United Kingdom.}
\address[icc]{Institute for Computational Cosmology, Durham University, South Road, Durham DH1 3LE, United Kingdom.}
\address[ggl]{Google Switzerland GmbH, Brandshenkestr. 110, 8002 Z\"urich, Switzerland.}

@@

Task-based parallelism is a conceptually simple paradigm for
shared-memory parallelism in which a computation is broken-down
into a set of inter-dependent tasks which are executed
concurrently.
Task dependencies are used to model the flow of data between
tasks, e.g.~if task $B$ requires some data generated by task $A$,
then task $B$ {\em depends} on task $A$ and cannot be executed
before task $A$ has completed.
The tasks and their dependencies can be seen as the nodes and edges
of a Directed Acyclic Graph (DAG) which can be
traversed in topological order, executing the tasks at the nodes
on the way down.

@@

In SMP superscalar \cite{ref:Perez2008}, StarPU \cite{ref:Augonnet2011},
QUARK \cite{ref:Yarkhan2011}, and KAAPI \cite{ref:Gautier2007}
the programmer specifies
what shared data each task will access, and how that data will
be accessed, e.g.~read, write, or read-write access.
The dependencies between tasks are then generated
automatically by the runtime system, assuming that the
data must be accessed and updated in the order in which
the tasks are generated.
StarPU also provides an interface for specifying additional
dependencies explicitly.
Intel's Threading Building Blocks (TBB)
\cite{ref:Reinders2010}
provide task-based parallelism using C++ templates in which
dependencies are handled either by explicitly waiting
for spawned tasks, or by explicitly manipulating
task reference counters.

@@

In all of these systems, the tasks are only aware of a single
type of relationship between each other, i.e. dependencies, which
specify a strict ordering between two tasks.
In many cases, however, the task ordering need not necessarily
be this strict.
Consider the case of two tasks that update some shared resource
in an order-independent way, e.g. when accumulating a result in
a shared variable, or exclusively writing to an output file.
In order to avoid concurrent access to that resource, it is
imperative that the execution of both tasks do not overlap,
yet the order in which the tasks are executed is irrelevant.
In the following, such a relationship will be referred to
as a {\em conflict} between two tasks.
\fig{TaskConflicts} shows a task graph with conflicting tasks
joined by thick dashed lines.
None of tasks $F$, $H$, and $I$ can be executed concurrently,
i.e. they must be serialized, yet in no particular order.

@@

This paper presents QuickSched, a framework for task-based
parallel programming with constraints, which aims to achieve
the following goals:
\begin{itemize}
    \item {\em Correct}: All constraints, i.e.~dependencies and
        conflicts, must be correctly enforced,
    \item {\em Fast}: The overheads associated with task management
        should be as small as possible,
    \item {\em Memory/cache-efficient}: Tasks accessing similar
        sets of data should be preferentially executed on the
        same core to preserve memory/cache locality as far as possible, and
    \item {\em Parallel-efficient}: Tasks should be executed in an order
        that sufficient work is available for all computational
        threads at all times.
\end{itemize}
\noindent 
Section~2 describes the main design considerations, Section~3 the
underlying algorithms and data structures, and
Section~4 their specific implementation in QuickSched.
Section~5 presents two test-cases:
\begin{enumerate}
    \item The tiled QR
    decomposition described in \cite{ref:Buttari2009} and for
    which the QUARK scheduler was originally developed, and
    \item A task-based Barnes-Hut tree-code to compute the
    gravitational N-body problem similar to the FMM codes
    of \cite{ref:Ltaief2012} and \cite{ref:Agullo2013},
\end{enumerate}
These real-world examples show how QuickSched can be used in practice,
and can be used to assess its efficiency.
Section~6 concludes with some general observations and future work
directions.

@@

From a programmer's perspective, there are two main paradigms for generating
task dependencies:
\begin{itemize}
  \item Implicitly via spawning and waiting, e.g.~as is done in Cilk, or
  \item Automatic extraction from data dependencies, e.g.~as is done in OmpSs.
\end{itemize}

The first scheme, spawning and waiting, is arguably the simplest to
use.
For simple depedency structures in which each task depends on only a
single task, i.e.~if the task DAG is a tree, each task {\em spawns}, or
creates, its dependent tasks after its completion (see \fig{Spawn}).
Hence for the tasks $A$--$E$ in \fig{Tasks}, task $A$ would spawn
tasks $B$ and $D$, task $B$ would spawn task $C$, and task $D$ would
spawn task $E$.
If a task has more than one dependency, e.g. tasks $D$--$F$ in \fig{Tasks},
then the task generation order is reversed: Task $E$ is executed first,
and first spawns tasks $D$ and $F$, and waits for both their completion
before doing its own computations.

@@

\begin{figure}
    \centerline{\epsfig{file=figures/Spawn.pdf,width=0.7\textwidth}}
    \caption{Two different task graphs and how they can be implemented
      using spawning and waiting.
      For the task graph on the left, each task spawns its dependent
      task or tasks. For the task graph on the right, in which task $C$
      has multiple dependencies, tasks $A$ and $B$ must be spawned and waited
      for by the calling thread before task $C$ can be spawned.}
    \label{fig:Spawn}
\end{figure}

Automatic dependency extraction, on the other hand,
works by enforcing dependencies
between tasks that access the same data.
These data dependencies are provided explicitly by the programmer, e.g.~by
describing which parameters to a task are input, output, and input/output.
The dependencies are enforced in the order in which the tasks are 
created.
This approach usually relies on compiler extensions, e.g.~ {\tt pragma}s
in C/C++, or a system of function call wrappers, to describe the task parameters
and their intent.

This approach allows programmers to specify rich dependency hierarchies
with very little effort, they still only allow for one type of relationship,
i.e.~dependencies, and lack the ability to deal with conflicts as
described in the previous section.
They may also not be able to understand more complex memory access patterns,
e.g.~for two tasks modifying the upper and lower triangular parts of a matrix,
which access the same block of memory, but never actually generate any
concurrency issues.

Here, QuickSched takes a drastically different approach by requiring
the programmer to create the complete task graph and its dependencies
explicitly before execution.
This approach has two main advantages:
\begin{itemize}
  \item It gives the user maximum flexibility with regards to the
    structure of the task graph generated,
  \item Knowing the structure of the task graph before execution
    allows the task scheduler to make more informed decisions
    as to how the tasks should be prioritized.
\end{itemize}

\noindent The obvious disadvantage is the burden of producing a correct
task graph is placed on the user.
Although some problems such as cyclic dependencies can be detected
automatically, there is no automatic way to detect whether the
dependencies actualy reflect the intent of the user.
Due to this added complexity, we consider QuickSched to be
a tool not designed for casual parallel programmers, but for 
those interested in investing more programming effort to achieve
better performance.

Conflicts between tasks or groups of tasks are not specified directly,
but are instead modeled as exclusive locks on a shared resource
which have to be obtained by a task before it can execute.
Thus, in \fig{TaskConflicts}, before executing, task $F$ has
to obtain an exclusive lock on the resource associated with
the conflict between tasks $F$, $H$, and $I$.
While task $F$ is being executed, neither $H$ nor $I$ can 
lock the same resource, and therefore will not execute until
task $F$ is done and the lock has been released.

The partitioning of the computation into tasks is also left entirely
to the user.
In theory, any program can be automatically converted to a task-based
representation since each statement in the program code
can be considered a single task, with dependencies to the
statements/tasks that produce the input values.
This set of basic tasks could, again in theory, be reduced by merging
tasks that share dependencies and/or resources using a variety
of graph algorithms.
The decomposition of a computation into tasks, however, usually
involves re-thinking the underlying algorithms such that they
best fit the task-based paradigm, e.g.~as in the examples in the
following sections, or as in \cite{ref:Gonnet2014}.
This process requires careful evaluation and is probably best
not left as an automatic transformation of an otherwise serial code.

Finally, the task granularity is an important issue: if the task
decomposition is too coarse, then good parallelism
and load-balancing will be difficult to achieve.
Converseley, if the tasks are too small, the costs of selecting and
scheduling tasks, which is usually constant per task, will
quickly destory any performance gains from parallelism.
Starting from a per-statement set of tasks, it is therefore
reasonable to group them by their dependencies and shared resources.

In the examples presented herein, we have chosen our task decomposition
and granularity such that
\begin{itemize}
  \item Each task maximizes the ratio of computation to data required,
  \item The resources required for each task fit comfortably in the
    lowest-level caches of the underlying system.
\end{itemize}
\noindent The first critera is biased towards bigger tasks, while the
second limits their size.
The criteria are thus used to optimize the
cache efficiency of the computation.

@@

The division of labor between the scheduler and
the queue objects is illustrated in \fig{QSched}.
The scheduler holds the tasks and is in charge
of managing {\em dependencies}.
Once a task has no unresolved dependencies, it is passed
on to a queue object.
The queue object, on the other hand, is in charge
of managing {\em conflicts}.
Computational threads can query a queue and will
receive only tasks for which all conflicts have been
resolved, i.e. for which all necessary resources could be 
exclusively locked.

There is also a division of responsibilities regarding
{\em efficiency} between the scheduler and the queue
objects.
The tasks in each queue are grouped according to the resources
they use, i.e. all the tasks in the same queue use a
similar set of resources.
The underlying assumption is that each computational
thread will preferentially access the same queue for tasks.
If the tasks in the queue share the same set of resources,
it increases the probability of said resources already
being present in the thread's cache, thus increasing
{\em cache efficiency}.
The scheduler is in charge of selecting the most appropriate
queue for each task, based on information stored in each task
on which resources are used.
Given a set of tasks with similar resources for which all
dependencies are resolved, the queue then decides which
tasks to prioritize.
This decision is made based on the length of the critical
path of the dependencies of each task.

@@

The {\tt unlocks} field points to the first element of
an array of {\tt nr\_unlocks} pointers to other tasks.
These pointers represent the dependencies in reverse:
if task $B$ depends on task $A$, then task $A$ {\em unlocks}
task $B$.
The unlocks therefore follow the direction of the arrows
in \figs{Tasks}{TaskConflicts}.
Conversely, {\tt wait} is the number of unresolved dependencies
associated with this task, i.e.~the number of unexecuted tasks
that unlock this task.
The wait-counters can
be set by initializing all the task wait counters to zero and then
incrementing the wait counter of each unlock-task of each task.

@@

Finally, {\tt cost} and {\tt weight} are measures
for the relative computational cost of this task, and the
relative cost of the critical path following the
dependencies of this task, respectively, i.e.~the task's cost
plus the maximum dependent task's weight (see \fig{TaskWeight}).
The task weights are computed by traversing
the tasks in reverse topological order following their dependencies, i.e.
as per \cite{ref:Kahn1962}, and computing each task's weight
up the task DAG.

\begin{figure}
    \centerline{\epsfig{file=figures/TaskWeight.pdf,height=0.3\textwidth}}
    \caption{Computation of the task weight.
      In this task graph, the height of each task corresponds to its
      computational {\em cost}.
      The {\em weight} of each task, e.g.~the topmost task,
      is computed from its cost plus
      the maximum dependent task's weight.
      This corresponds to the computatoinal cost of the critical
      path following the task's dependencies.}
    \label{fig:TaskWeight}
\end{figure}

@@

The {\tt lock} field is either {\tt 0} or {\tt 1} and indicates
whether this resource is currently in use, i.e.~{\em locked}.
To avoid race conditions, this value should be tested
and set using atomic instructions only.
The {\tt hold} field is a counter indicating how many
sub-resources of the current resource are locked.
If a resource's hold counter is not zero, then it is
{\em held} and cannot be locked.
Likewise, if a resource is locked, it cannot be held
(see \fig{Resources}).

@@

Incrementing the hold counter of the resource can be implemented
as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void resource_hold(struct resource *r) {
  if (atomic_cas(&r->lock, 0, 1) != 0) return 0;
  atomic_inc(&r->hold);
  r->lock = 0;
  return 1;
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt atomic\_cas(val,old,new)} is an atomic
compare-and-swap operation that sets {\tt val} to {\tt new}
if it is currently equal to {\tt old}.
Similarly, {\tt atomic\_inc(val)} increments {\tt val} by one
atomically.
The resource's {\tt lock} is used to check if the resource
is already locked (line 2), and it is held while {\tt hold}
is incremented, to avoid overlapping hold/lock operations.
If the resource can be locked, the hold counter is incremented
atomically (line~3), and the lock is released (line~4),
returning {\tt 1} or {\tt 0} if the resource could be held
or not, respectively.

The locking procedure itself is implemented as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void resource_lock(struct resource *r) {
  struct resource *up, *top;
  if (r->hold && atomic_cas(&r->lock, 0, 1) != 0)
    return 0;
  if (r->hold) {
    r->lock = 0;
    return 0;
  }
  for (up = r->parent; up != NULL; up = up->parent)
    if (!resource_hold(up)) break;
  if ((top = up) != NULL) {
    for (up = r->parent; up != top, up = up->parent)
      atomic_dec(&up->hold);
    r->lock = 0;
    return 0;
  } else
    return 1;
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where in line~3 the resource is first locked if it
is not held.
Due to a possible race condition when holding the resource,
the {\tt hold} counter must be checked again once the resource
has been locked (line~5).
In lines~9--10 the hold counters of the hierarchical parents
are incremented using the procedure described earlier.
If this process fails at any point (line~11), the
previously set hold counters are decremented (line~12)
and the lock is released (line~13).
The procedure then returns {\tt 1} or {\tt 0} if the resource
could be locked or not, respectively.

Finally, unlocking a resource is relatively straight-forward,
i.e.~the lock is set to zero and the hold counters up the
resource hierarchy are decremented.

@@

One possible strategy would be to maintain an array of tasks
sorted by their weights, and to traverse that list in descending
order, trying to lock the resources of each task, until
a lockable task is found, or returning a failure otherwise.
Although this would return the best possible task, it
requires maintaining a sorted array in which inserting
or removing an entry is in \oh{n} for $n$ elements.
Using an unsorted array would require only \oh{1} operations for
insertion and deletion, but is undesirable as it completely
ignores the task weights.

@@

The data structure for the queue is defined as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct queue {
  struct task **tasks;
  int count, lock;
};
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt tasks} is an array of {\tt count} pointers
to the tasks in max-heap order, and {\tt lock} is used to
guarantee exclusive access to the queue.

Inserting a task in the queue is relatively straight-forward:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void queue_put(struct queue *q, struct task *t) {
  while (atomic_cas(q->lock, 0, 1) != 0) {}
  q->tasks[q->count++] = t;
  bubble-up the q->count - 1st entry of q->tasks.
  q->lock = 0;
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the loop in line~2 spins until an exclusive
lock on the queue can be obtained.
The task is added to the end of the heap array (line~3)
and the heap order is fixed (line~4).
Before returning, the lock on the queue is released (line~5).

Obtaining a task from the queue can be implemented as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *queue_get(struct queue *q) {
  struct task *res = NULL;
  int j, k;
  while (atomic_cas(q->lock, 0, 1) != 0) {}
  for (k = 0; k < q->count; k++) {
    for (j = 0; j < q->tasks[k]->nr_locks; j++)
      if (!resource_lock(q->tasks[k]->lock[j])) break;
    if (j < q->tasks[k]->nr_locks)
      for (j = j - 1; j >= 0; j--)
        resource_unlock(q->tasks[k]->lock[j]);
    else
      break;
  }
  if (k < q->count) {
    res = q->tasks[k];
    q->tasks[k] = q->tasks[--q->count];
    trickle-down the kth entry of q->tasks.
  }
  q->lock = 0;
  return res;
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where as with the queue insertion, the queue is first
locked for exclusive access (line~4).
The array of task pointers is then traversed (line~5),
locking the resources of each task (lines~6--7).
If any of these locks fail (line~8), the locks that were obtained
are released (lines~9--10), otherwise, the traversal is aborted
(line~12).
If all the locks on a task could be obtained (line~14), the
task pointer is replaced by the last pointer in the heap (line~16)
and the heap order is restored (line~17).
Finally, the queue lock is released (line~19) and the locked task
or, if no lockable task could be found, {\tt NULL} is returned.

@@

This section presents two test cases showing
how QuickSched can be used in real-world applications, and
providing benchmarks to assess its efficiency and scalability.
The first test is the tiled QR decomposition originally
described in \cite{ref:Buttari2009}, which has been used as a benchmark
by other authors \cite{ref:Agullo2009b,ref:Badia2009,ref:Bosilca2012}.
This example only requires dependencies and is presented 
as a point of comparison to existing task-based parallel
programming infrastructures.
The second example is a Barnes-Hut tree-code, a problem
similar to the Fast Multipole Method described in both
\cite{ref:Ltaief2012} and \cite{ref:Agullo2013}.
This example shows how conflicts, modeled
via hierarchical resources, can be useful in modelling and executing
a problem efficiently.

The source code of both examples is distributed with the
QuickSched library, along with scripts to run the benchmarks
and generate the plots used in the following.
All examples were compiled with gcc v.\,4.8.2 using the
{\tt -O2 -march=native} flags and run on
a 64-core AMD Opteron 6376 machine running
at 2.67\,GHz.
